INGRID: Creating Languages in MPS from ANTLR Grammars

### 2017-06-13 chapters/background.tex
making changes

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}
\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (Spoofax, apod).}

@@

JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
The code can then be transformed into a target language, typically a GPL, such as Java or C, and eventually compiled into executable programs.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since \todo{programs in these languages} are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

Besides that, we are aware of three other projects that provide certain support for GPLs in the context of MPS --- namely PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
We describe their main features and limitations in this section.
\todo{Check what has been done for Spoofax.}

@@

\todo{Najit dalsi podobne systemy (treba Spoofax) a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}

@@

\todo{
Mention other state-of-the-art language workbenches: Spoofax.
	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010
	Eclipse-based IDE support for SDF-based languages.
	Parser-based language workbench.
		Syntax essentially restricted to linear text.
		But it supports language extension and composition.
}

>>>>

\todo{For each MPS term (Structure, Editor, TextGen, concept, and so on) say which generally/widely known terms (from related academic work) are similar/analogous. Relate the MPS-specific terms to the large body of academic work.}
\todo{Mozna rict jak se zakladni aspekty (Structure, Editor, TextGen) nazyvaji v jinych IDE (MetaCase MWB, apod).}

@@

JetBrains MPS is a language workbench --- an IDE that allows developers to create their own languages and use them to write code.
The code can then be transformed into a target language, typically a GPPL, such as Java or C, and eventually compiled into executable programs.

@@

\paragraph{TextGen.}
The TextGen aspect specifies how a given AST node will be translated into plain text representation.
It is typically needed only for the bottom-line base languages.
DSLs, on the other hand, need to define rules for model-to-model conversions (Generators), since \todo{programs in these languages} are rarely converted to text directly.
After TextGen has generated textual sources from an AST, a compiler for the particular GPPL is invoked to compile the textual sources into binary code.
The TextGen definition follows a very straightforward pattern --- each node outputs its textual representation into a buffer, while calling TextGen of its children nodes at the right moments.
MPS calls the corresponding method on the root AST nodes of the given program.

@@

Besides that, we are aware of three other projects that provide certain support for GPPLs within projectional editors, however all of them target MPS.
The projects are named PE4MPS, ANTLR{\_}MPS, and mps-metabnf.
We describe their main features and limitations in this section.

@@

\todo{Najit dalsi podobne systemy a kratce popsat existujici reseni podobnych uloh (automated import of general-purpose languages) v jejich kontextu.}

@@

\todo{
\paragraph{Other IDEs.}
There also exist several language workbenches that are based on parser generators (such as ANTLR) and allow free editing of textual source code (unlike the projectional editor in MPS).
A prominent example/tooo in this category is Spoofax [ref].
	Reference: L.C.L. Kats and E. Visser. The Spoofax Language Workbench: Rules for Declarative Specification of Languages and IDEs. OOPSLA 2010, ACM.
	Syntax of DSLs is essentially restricted to linear text.
INGRID is not applicable to such tool/workbenches (for creating languages inside them).
}

### 2017-06-13 chapters/introduction.tex
making changes

An alternative approach is projectional (or structural) editing~\cite{ref:VWK15} \todo{Pridat citaci: nedavno precteny clanek o srovnani projectional editors}, where a developer directly manipulates the AST representation of the source code instead of plain text.
This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax \todo{Pridat referenci}.
In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
We provide more technical details about MPS in Section~\ref{sect:MPS}.
In the rest of the paper, we will often use \todo{examples from} MPS to illustrate the key concepts of our approach and related issues.

@@

Projectional editing, keeping the code in the AST form, and the absence of parsers, brings along several benefits.
\begin{itemize}
	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
		%An AST created by the editor can represent only a syntactically valid code.
	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
	\item The languages may support contextual or non-parseable notations, such as tables, diagrams and positional syntax. \todo{Zminit layout.}
	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
\end{itemize}
All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
\todo{Popsat dalsi benefits of projectional editors/IDEs (and of MPS in particular): Projectional editors never have to parse the concrete syntax, so they can support/use more complicated syntax/layout. Support for non-linear notations (mathematical expressions, graphical or tabular layout), notations that use two-dimensional layout. Projectional editors mix different notations/styles.}
\todo{Nekam pridat text: Benefits and limitations of projectional editors are thoroughly discussed in [M. Voelter, J. Siegmund, T. Berger, and B. Kolb. Towards User-Friendly Projectional Editors. SLE 2014, LNCS 8706].}

On the other hand, the usage of projectional editors introduces some new problems.
Before a language can be used inside editors\todo{/IDEs} such as MPS, it has to be defined through the \todo{corresponding/specific} infrastructure, so that the IDE can understand the language and work with it.
Specifically, an author of a language has to create:
\begin{enumerate}
	\item the abstract syntax (structure of the language), which defines the types of allowed AST nodes,
	\item the concrete syntax used for editing, i.e. projection of the AST on the screen and its interactions with the user, and
	\item text generation scripts to enable creation of plain text files used as input for compilers.
\end{enumerate}

@@

\todo{IDE systems based on projectional editors are} typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPL), such as C or Java.
This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
Therefore, the target GPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPL.

However, very few mainstream GPLs are now fully supported by MPS \todo{Zjistit jako to maji dalsi language workbenches (Spoofax, apod)}, because it requires substantial effort to manually create a full definition of a GPL in MPS.
Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPLs.
The overall goal of our project is to automatize the process of language definition in MPS as much as possible, exploiting the fact that a big part of this process is quite straightforward.
We believe that such an automated process would encourage and speed-up the migration of more GPLs into MPS, thus giving the authors of DSLs more options.
Other possible applications include the support for writing programs in GPLs using the projectional editor and combination of GPLs with domain-specific languages natively created in MPS.

@@

\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches based on projectional editing.}
\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges and principles behind our approach are general and they apply also to other language workbenches.
\todo{Although we present INGRID on MPS, we believe it can be easily adapted to other language workbenches.}

\todo{Zduraznit (vyzdvihnout) general research contribution (applicable to MPS and other platforms).
	Mozna rict neco jako "INGRID efektivne resi problem XY".
	We summarize/highlight our contribution:
	\begin{itemize}
		\item Metoda pro konstrukci language definitions usable/suitable for IDEs based on projectional editing, in particular for JetBrains MPS, from ANTLR grammars, ktera je pouzitelna pro (1) import of general-purpose programming languages a (2) efficient construction of DSLs. To our belief, the method can be easily adapted to other language workbenches (e.g., Spoofax).
		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
	\end{itemize}
}

The rest of the paper has the following structure.
In the next section, we provide necessary information about MPS and a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.

>>>>

An alternative approach is projectional (or structural) editing~\cite{ref:VWK15,ref:VSB14}, where a developer directly manipulates the AST representation of the source code instead of plain text.
This idea emerged as early as in 1970s, but it failed to get adopted widely, mostly due to inconvenient and unnatural way of manipulating code.
A recent revival of projectional editing has been observed in the area of language workbenches --- IDE-like tools that enable the developers to manipulate the actual language definition.
Popular IDEs (examples) include JetBrains MPS~\cite{ref:MPS,ref:MPSBOOK} and Spoofax~\cite{ref:KV10}.
In particular, JetBrains MPS (Meta Programming System) is an open-source language workbench that focuses on domain specific languages (DSL) and leverages the concept of projectional editing.
MPS provides the whole IDE infrastructure that enables developers to design custom languages, use them to write program code, and compile the code into executables.
%We provide more technical details about MPS in Section~\ref{sect:MPS}.
In the rest of the paper, we will often use examples from MPS to illustrate the key concepts of our approach and related issues.

@@

Projectional editing, keeping the code in the AST form, and the absence of parsers, brings along several benefits.
\begin{itemize}
	\item The developers practically cannot enter syntactically invalid code, because a projectional editor controls the interaction between the user and the code.
		%An AST created by the editor can represent only a syntactically valid code.
	\item Programming languages can be defined in a modular way, and multiple languages can be easily combined together in a single program or one can extend another, while parsers seriously limit language modularity for traditional parsed languages.
	\item The languages may support diverse contextual or non-parseable notations with complex layout, such as tables, mathematical expressions, diagrams and positional syntax.
	\item Since the projection is detached from the physical representation of code (AST), authors of languages can define multiple notations and allow the developers to switch between different representations of the code on the screen.
\end{itemize}
All of this is useful especially for DSLs, which are frequently used by domain experts who may not be professional software developers.
The mbeddr project~\cite{ref:MBEDDR}, which extends C with numerous domain-specific constructs and data types, illustrates the abilities of language workbenches in general and of JetBrains MPS in particular.
Voelter et al.~\cite{ref:VSB14} discuss the benefits and limitations of projectional editors.

Nevertheless, the usage of projectional editors introduces some new problems.
Before a language can be used inside editors\todo{/IDEs} such as MPS, it has to be defined through the \todo{corresponding/specific} infrastructure, so that the IDE can understand the language and work with it.
Specifically, an author of a language has to create:
\begin{enumerate}
	\item the abstract syntax (structure of the language), which defines the types of allowed AST nodes,
	\item the concrete syntax used for editing, i.e. projection of the AST on the screen and its interactions with the user, and
	\item text generation scripts to enable creation of plain text files used as input for compilers.
\end{enumerate}

@@

\todo{IDE systems based on projectional editors are} typically used for syntactically rich DSLs, which are likely to benefit most from overcoming the limitations of parser-based languages.
Before a program written in a rich DSL can be executed, it is \todo{has to be} transformed on the AST level by a series of model-to-model transformations to an AST model that represents the desired semantics in a general-purpose language (GPPL), such as C or Java.
This model is then converted to a textual representation of that language and compiled by the standard means of the target platform.
Therefore, the target GPPL must also be defined by the MPS means in order to allow the DSLs to have their models transformed into a model in the GPPL.

However, very few mainstream GPPLs are now fully supported by MPS \todo{(and other language workbenches based on the same principles, such as MetaCase MetaEdit+ WorkBench (MWB) \url{http://www.metacase.com/mwb/})}, because it requires substantial effort to manually create a full definition of a GPPL in MPS.
Only Java and C have been implemented to date, which limits the authors of DSLs to only target these supported GPPLs.
The overall goal of our project is to automatize the process of language definition in MPS as much as possible, exploiting the fact that a big part of this process is quite straightforward.
We believe that such an automated process would encourage and speed-up the migration of more GPPLs into MPS, thus giving the authors of DSLs more options.
Other possible applications include the support for writing programs in GPPLs using the projectional editor and combination of GPPLs with domain-specific languages natively created in MPS.

@@

\todo{Although we define the INGRID method in the context of MPS, it can be easily adapted for/towards construction of language definitions usable in (suitable for) other IDEs and language workbenches (i) based on projectional editing (which let the user edit the AST representation (abstract syntax) directly) and (ii) where language definition is based on structural object-oriented approach.}
\todo{Zmenit: we explain our tool using examples from MPS and ANTLR} Although we explain everything on the examples of MPS languages and ANTLR grammars, we emphasize that most of the challenges and principles behind our approach are general and they apply also to other language workbenches.
\todo{Although we present INGRID on MPS, we believe it can be easily adapted to other language workbenches.}

\todo{Zduraznit (vyzdvihnout) general research contribution (applicable to MPS and other platforms).
	Mozna rict neco jako "INGRID efektivne resi problem XY".
	We summarize/highlight our contribution:
	\begin{itemize}
		\item Metoda pro konstrukci language definitions usable/suitable for IDEs based on projectional editing, in particular for JetBrains MPS, from ANTLR grammars, ktera je pouzitelna pro (1) import of general-purpose programming languages a (2) efficient construction of DSLs. To our belief, the method can be easily adapted to other language workbenches based on similar principles as MPS.
		\item Discussion of general challenges that apply to differences between (1) the definition of a programming language that uses grammars in an EBNF-like notation and (2) the structured object-oriented approach used in projectional editors.
		\item We discuss how practical usability of generated MPS languages depends on the specific definition of ANTLR grammar and its manual adjustments that improve usability (details in Section 3.6).
	\end{itemize}
}

The rest of the paper has the following structure.
In the next section, we provide (i) all information about MPS that is necessary to understand the examples from MPS presented in the paper and (ii) a brief overview of related work, including projects with similar goals as \textsc{Ingrid}.
Then, in Section~\ref{sect:LANGDEF}, we describe our approach to the construction of a language definition in MPS from its ANTLR grammar, discussing the major challenges together with our solution along the way.
We also discuss our experience with usage of \textsc{Ingrid} on complex mainstream languages, such as JavaScript and C\#, and highlight open problems (Section~\ref{sect:EVAL}), and then we conclude.
